Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-03-31 paper/manuscript.tex
Remove unnecessary whitespace at line endings
Move figures to appear on page that they are mentioned in text
some minor tweaks

\begin{abstract}
% no more than 250 words.
Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
The evolutionary success of influenza virus lineages depends on how these mutations affect HA's functionality and antigenicity.
Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain.
We show that mutations that are measured to be more favorable in cell culture tend to result in more evolutionarily successful H3N2 viral lineages than mutations that are less favorable in cell culture.
Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding evolution in nature.
We also compare our measurements for H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
For instance, the H3 HA has less disparity in mutational tolerance between the head and stalk domains than the H1 HA.
Overall, our work suggests that experimental measurements of mutational effects can be leveraged to help understand the evolutionary fates of viral lineages in nature --- but only when the measurements are made on a viral strain similar to the ones being studied in nature.
\end{abstract}

@@

\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein~\cite{fitch1997long, bhatt2011genomic}.
Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping}.
This evolution is further characterized by competition and turnover among groups of strains called clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
Clades vary widely in their evolutionary success with some dieing out soon after emergence and others going on to take over the virus population.
Several lines of evidence indicate that successful clades, which reach high frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.

@@

It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
Here, we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
We then show that these experimental measurements can help discriminate evolutionarily successful mutations from those that quickly die out.
However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.

@@

@@

\begin{table}
\caption{\label{tab:phydms}
{\bf Substitution models informed by the experiments describe H3N2's natural evolution better than traditional substitution models.}}
\begin{center} 
\begin{tabular}{cccccccc}
\hline
\bf{Model} & \bf{$\Delta$AIC} & \bf{LnL} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
ExpCM & 0.0 & -8441 & 2.47 & 0.91 \\
GY94 M5 & 2094 & -9482 & -- & 0.36 (0.30, 0.84) \\
ExpCM, site avg. & 2501 & -9692 & 0.67 & 0.32 \\
GY94 M0 & 2536 & -9704 & -- & 0.31 \\
\hline
\end{tabular}
 \end{center}
\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
For the M5 model, we give the mean followed by the shape and rate parameters of the gamma distribution over $\omega$.
}
\end{table}

@@

\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
Our experiments measure which amino acids are tolerated at each HA site under selection for protein function.
We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.

We performed a similar analysis using our new data for the Perth/2009 H3 HA.
Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B). 
This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
Helix A is a component of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.

@@

\subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
Mutations occurring in the H3N2 virus population have widely varying evolutionary outcomes (Figure~\ref{fig:frequency_trajectory}).
Some mutations appear, spread and fix in the population, while others briefly circulate before dieing out.
We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
If a mutation is highly deleterious, we would expect it to quickly die out, whereas less deleterious or beneficial mutations are more likely to expand in frequency.
Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
As such, each mutation is shown as a separate circle occuring on a separate branch in Figure \ref{fig:frequency_trajectory}.
However, because multiple mutations on the same phylogeny branch cannot be disentangled, we identified all mutations on a branch and assigned a single mutational effect based on the sum of effects for each mutation on the branch.
We observed few circulating mutations in our phylogeny with strongly deleterious mutational effects and rarely observed these mutations reaching high frequency (Figure~\ref{fig:frequency_trajectory}).
Mutations that reached high global frequencies generally had neutral or beneficial mutational effects.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.png}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally determined effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The bottom panel shows the frequency trajectory of each mutation, with mutation trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
}
\end{figure}

@@

\subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
An obvious hypothesis for why the H1 deep mutational scanning is not useful for understanding the evolution of H3N2 influenza viruses is that the effect of the same mutation is often different between these two HA subtypes.
To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.

To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2017mapping}.
Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
Although some sites have small shifts near zero, many sites have large shifts.
These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the typical shift between H3 and H1 is still smaller than that observed when comparing HA to the non-homologous HIV Env protein. 
Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
\caption{\label{fig:distance_distribution}
{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled. 
These HAs have 42\% amino-acid identity.
(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray. 
$R$ indicates the Pearson correlation coefficient.
(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
However, the shifts between H3 and H1 are usually less than the differences between HA and HIV Env (green).
}
\end{SCfigure*}

@@

We measured the effects of all single amino-acid mutations to a specific HA, and then generalized these measurements to other H3N2 HAs from a 50-year timespan.
These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}, and the extent to which this occurs is an important question in its own right.
Our work suggests that measurements on a HA from single human H3N2 viral strain can be usefully generalized to at least some extent across the entire evolutionary history of human H3N2 HA.
On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2017mapping}.
Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.

@@

\matmethods{
\subsection*{Data and computer code}
Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
Computer code used to analyze the data and produce the results in the paper are on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.

@@

\subsection*{Analysis of deep sequencing data}
We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.0}) to analyze the deep sequencing data.
The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
Computer code that performs the entirety of this analysis is available on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.

@@

\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid}, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).

@@

\subsection*{Analysis of mutational shifts}
To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.

>>>>

\begin{abstract}
% no more than 250 words.
Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
The evolutionary success of influenza virus lineages depends on how these mutations affect HA's functionality and antigenicity.
Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain.
We show that mutations that are measured to be more favorable for viral growth are enriched in evolutionarily successful H3N2 viral lineages relative to mutations that are measured to be less favorable for viral growth.
Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding evolution in nature.
We also compare our measurements for H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
For instance, the H3 HA has less disparity in mutational tolerance between the head and stalk domains than the H1 HA.
Overall, our work suggests that experimental measurements of mutational effects can be leveraged to help understand the evolutionary fates of viral lineages in nature --- but only when the measurements are made on a viral strain similar to the ones being studied in nature.
\end{abstract}

@@

\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein~\cite{fitch1997long, bhatt2011genomic}.
Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping}.
This evolution is further characterized by competition and turnover among groups of strains called clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
Clades vary widely in their evolutionary success with some dying out soon after emergence and others going on to take over the virus population.
Several lines of evidence indicate that successful clades, which reach high frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.

@@

It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high,haddox2018mapping}.
However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
Here, we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
We then show that these experimental measurements can help discriminate evolutionarily successful mutations from those that are found in strains that quickly die out.
However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.

@@

\begin{table}[b!]
\caption{\label{tab:phydms}
{\bf Substitution models informed by the experiments describe H3N2's natural evolution better than traditional substitution models.}}
\begin{center}
\begin{tabular}{cccccccc}
\hline
\bf{Model} & \bf{$\Delta$AIC} & \bf{LnL} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
ExpCM & 0.0 & -8441 & 2.47 & 0.91 \\
GY94 M5 & 2094 & -9482 & -- & 0.36 (0.30, 0.84) \\
ExpCM, site avg. & 2501 & -9692 & 0.67 & 0.32 \\
GY94 M0 & 2536 & -9704 & -- & 0.31 \\
\hline
\end{tabular}
 \end{center}
\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
For the M5 model, we give the mean followed by the shape and rate parameters of the gamma distribution over $\omega$.
}
\end{table}

@@

@@

\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
Our experiments measure which amino acids are tolerated at each HA site under selection for viral growth.
We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.

We performed a similar analysis using our new data for the Perth/2009 H3 HA.
Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.png}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally determined effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The bottom panel shows the frequency trajectory of each mutation, with mutation trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
}
\end{figure}

\subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
Mutations occurring in the H3N2 virus population have widely varying evolutionary outcomes (Figure~\ref{fig:frequency_trajectory}).
Some mutations appear, spread and fix in the population, while others briefly circulate before dying out.
We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
If a mutation is highly deleterious, we would expect it to quickly die out, whereas less deleterious or beneficial mutations are more likely to expand in frequency.
Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
As such, each mutation is shown as a separate circle occurring on a separate branch in Figure \ref{fig:frequency_trajectory}.
However, because multiple mutations on the same phylogeny branch cannot be disentangled, when multiple mutations occurred on a single branch, we assigned a single mutational effect based on the sum of effects of each mutation.

After annotating mutations and their frequencies on the phylogeny in this way (Figure~\ref{fig:frequency_trajectory}), it is visually obvious that there are relatively few circulating mutations with strongly deleterious mutational effects---and that such deleterious mutations rarely reach high frequency when they do occur.
Mutations that reached high global frequencies generally had neutral or beneficial mutational effects.

@@

\subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
An obvious hypothesis for why the H1 deep mutational scanning is not useful for understanding the evolution of H3N2 influenza viruses is that the effect of the same mutation is often different between these two HA subtypes.
To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2018mapping}.
However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.

To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2018mapping}.
Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
Although some sites have small shifts near zero, many sites have large shifts.
These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2018mapping}.
However, the typical shift between H3 and H1 is still smaller than that observed when comparing HA to the non-homologous HIV Env protein.
Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
\caption{\label{fig:distance_distribution}
{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled.
These HAs have 42\% amino-acid identity.
(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray.
$R$ indicates the Pearson correlation coefficient.
(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2018mapping}, and plotted the distribution of shifts for all sites.
The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
However, the shifts between H3 and H1 are usually less than the differences between HA and HIV Env (green).
}
\end{SCfigure*}

@@

We measured the effects of all single amino-acid mutations to a specific HA, and then generalized these measurements to other H3N2 HAs from a 50-year timespan.
These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2018mapping}, and the extent to which this occurs is an important question in its own right.
Our work suggests that measurements on a HA from single human H3N2 viral strain can be usefully generalized to at least some extent across the entire evolutionary history of human H3N2 HA.
On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2018mapping}.
Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.

@@

\matmethods{
\subsection*{Data and computer code}
Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
Computer code used to analyze the data and produce the results in the paper are on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.

@@

\subsection*{Analysis of deep sequencing data}
We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.0}) to analyze the deep sequencing data.
The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
Computer code that performs the entirety of this analysis is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript} \comment{Juhye, add additional details and link to the exact notebook}.

@@

\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid}, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript} \comment{Juhye: add link to exact file}.
The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).

@@

\subsection*{Analysis of mutational shifts}
To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2018mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
